---
title: "7-Multivariate-analysis"
output: html_document
date: '2022-05-12'
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 7.1 Revise SVD
Redo the analysis with a rank 2 matrix

```{r}
n <- 100
p <- 4
Y2 <- outer(rnorm(n), rnorm(p)) + outer(rnorm(n), rnorm(p))

head(Y2)

ggplot(data.frame(Y2), aes(x = X1, y=X2, col=X3, size=X4)) + geom_point()
```
Now there are obviously at least two dimensions because if we project the data onto the first two coordinates (by default called X1 and X2 when you convert a matrix into a data frame in R), then the data varies in both dimensions. So the next step is to try to decide if there are more than two dimensions. The top right points are the closest to you (they’re biggest) and as you go down and left in the plot those points are farther away. In the left are the bluest points and they seem to get darker linearly as you move right.

```{r}
svd(Y2)$d
Y = Y2 + matrix(rnorm(n*p, sd=0.01), n, p)
svd(Y)$d
```

## 7.2
7.2a: create a first a matrix of highly correlated bivariate data such as that shown in Figure 7.35.
Hint: Use the function mvrnorm.

Check the rank of the matrix by looking at its singular values.

7.2b: perform a PCA and show the rotated principal component axes.

```{r}
#7.2a
mu1 = 1; mu2 = 2; s1 = 2.5; s2=0.8; rho = 0.9;
sigma = matrix(c(s1^2, s1*s2*rho, s1*s2*rho, s2^2), 2)

library('MASS')
sim2d <- data.frame(mvrnorm(50, mu = c(mu1, mu2), Sigma = sigma))

svd(scale(sim2d))$d
svd(scale(sim2d))$v[, 1]

ggplot(data.frame(sim2d),aes(x=X1,y=X2)) +
    geom_point()

# 7.2b
respc=princomp(sim2d)
dfpc = data.frame(pc1=respc$scores[,1],
pc2 = respc$scores[,2])
 ggplot(dfpc,aes(x=pc1,y=pc2)) +
   geom_point() + coord_fixed(2)
```

## 7.4
Let’s revisit the Hiiragi data and compare the weighted and unweighted approaches.

7.4a: make a correlation circle for the unweighted Hiiragi data xwt. Which genes have the best projections on the first principal plane (best approximation)?

```{r}
library("Hiiragi2013")
library("ade4")
library("factoextra")
library("pander")
library("knitr")
library("tidyverse")
library("ggrepel")

data('x', package='Hiiragi2013')
xwt <- x[, x$genotype == 'WT']
sel <- order(rowVars(Biobase::exprs(xwt)),
             decreasing=T)[1:100]
xwt <- xwt[sel,]
head(t(Biobase::exprs(xwt)))
```

Angles between vectors: measure of correlation btw the 2 corresp. genes.

length of arrow: correlation of a gene w/ 1st principal axes

1st PC: linear comb. of genes w/ max variance

2nd PC: linear comb. of genes w/ max variance orthogonal to 1st PC

```{r}
# no row.w 
pca.xwt <- dudi.pca(as.data.frame(t(Biobase::exprs(xwt))),
                    center = T, scale = T, nf = 2, scannf = F)

fviz_pca_var(pca.xwt, col.circle = 'black', geom = 'arrow') + ggtitle('')

# weighted
tab <- table(xwt$sampleGroup)
xwt$weight = 1 / as.numeric(tab[xwt$sampleGroup])
pca.xwt.weight <- dudi.pca(as.data.frame(t(Biobase::exprs(xwt))),
                    row.w = xwt$weight,
                    center = TRUE, scale = TRUE, nf = 2, scannf = FALSE)
fviz_pca_var(pca.xwt.weight, col.circle = "black",geom= "arrow") + ggtitle("") 
```

Plot genes with 'long' arrow
[Soure](https://kind-neumann-789611.netlify.app/post/exercise-solution-for-chapter-7/)
```{r}
corrCircle <- fviz_pca_var(pca.xwt, col.circle = "black")$data
arrowLengths <- sqrt(corrCircle$x^2+corrCircle$y^2)
cutoff <- 0.8
kpInd <- order(arrowLengths, decreasing=TRUE)[1:sum(arrowLengths>cutoff)]
genes <- corrCircle[kpInd,"name"]
genes

corrCircle %>% 
  mutate(length = sqrt(x^2 + y^2)) %>% 
  dplyr::filter(length >= 0.8) %>% 
  ggplot(aes(x = 0, xend = x, y = 0, yend = y)) + 
  geom_segment(arrow = arrow(length = unit(0.1, "inches"))) + 
  geom_label_repel(aes(x = x, y = y, label = name), 
                   size = 2, alpha = 0.7) + 
  coord_fixed()
```

7.4b: make a biplot showing the labels of the extreme gene-variables that explain most of the variance in the first plane. Add the the sample-points.

```{r}
top_genes <- rep('low', dim(xwt)[1])
top_genes[1:100 %in% kpInd] <- 'top'
fviz_pca_biplot(pca.xwt, col.var = top_genes,
               col.ind = xwt$sampleGroup, 
               label = '') +
  ggtitle('')

fviz_pca_biplot(pca.xwt.weight, col.var = top_genes,
               col.ind = xwt$sampleGroup, 
               label = '') +
  ggtitle('')

```

colors of arrows: whether corresp. gene had length >/< 0.8 on the corrcircle

Unweighted: EPI & PE groups (w/ least no of samples) located further away than the rest fr 1st principle
Weighted: Not the case as all groups are already weighted uniformly
